Polynomial Time and Full Dependent Types

### 2023-07-11 paper/paper.tex
[ paper ] Complete programming bit, start on bibliography

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Robert Atkey}
\email{robert.atkey@strath.ac.uk}
\orcid{1234-5678-9012}
\affiliation{%
  \institution{University of Strathclyde}
  \streetaddress{26 Richmond Street}
  \city{Glasgow}
  \country{UK}
  \postcode{G1 1XH}
}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We combine dependent types with linear type systems that soundly and
  completely capture polynomial time computation. To combine dependent
  and linear types, we use \emph{Quantitative Type Theory}. We explore
  two systems for capturing polynomial time: one system that disallows
  construction of iterable data, and one, based on the LFPL system of
  Martin Hofmann, that controls construction via a payment
  method. Both of these are extend to full dependent types, allowing
  for arbitrary computation in types alongside guaranteed polynomial
  time computation in terms. We prove the soundness of the systems
  using a realisability technique due to Hofmann and Dal Lago.

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{type theory, implicit computational complexity, linear logic, computational complexity}

@@

Typical Type Theory only speaks of the public face that programs
present to the world -- if you input things like this, you get things
that look like that -- but cannot bring itself to mention the true
cost of programs' execution. One can encode costs by embedding another
programming language in type theory \cite{pottier}, or one can
synthesise costs by treating resource counting as a computational
effect \cite{daniellsson,calf}, but neither of these capture the
intrinsic costs of the programs we write in type theory. These
techniques deliver only \emph{conspicuous consumption}, not speaking
of the real resources consumed.

@@

This paper makes the following contributions to the theory and use of
linear dependent type theory and implicit computational complexity:
\begin{enumerate}
\item We formulate two systems that combine linear type theory for
  polytime computation with full dependent types, using Quantitative
  Type Theory. The system is presented in \autoref{sec:qtt}. The
  linear typing discipline required for enforcing polytime is provided
  by QTT, but we also need to carefully add constructs for
  non-iterable datatypes (\autoref{sec:noniter-qtt}) and the two kinds
  of natural number iterator that we consider
  (\autoref{sec:cons-free-qtt} and \autoref{sec:lfpl-qtt}). Porting
  the natural number iterators from the simply typed to the
  dependently typed setting requires careful annotation of the rules
  to ensure that the correct information is available for type
  checking, while also not allowing too much information to be made
  available at runtime that would violate the polytime soundness
  property. A further contribution of this paper is the addition of
  reflection types to QTT, \autoref{sec:qtt-reflection}, which allow
  statements about polytime realisability to be reflected into
  types. We explain in \autoref{Sec:realisability-future} how this
  will have applications in QTT beyond the polytime systems presented
  here.
\item We demonstrate the utility of the combination of polytime and
  dependent types in \autoref{sec:programming-polytime}. Just as in
  the simply typed world, we have an expressive language for writing
  polytime programs. With the additional power of dependent types, we
  can also prove properties of these programs. A simple example is
  proving that a sorting program actually sorts. Using QTT reflection,
  we can go further and represent the class of polytime problems, with
  polytime reductions between them, as dependent pairs
  (\autoref{sec:polytime-problems}). Our final examples use dependent
  types to give monadic presentations of the complexity classes of
  Non-deterministic Polynomial (NP) time and Probabilistic Polynomial
  (PP) time. Since these classes rely on specific correctness
  criteria, it is not possible to capture them in a simply typed
  system for polytime.
\item We prove the polytime soundness of our systems via a
  realisability argument in \autoref{sec:soundness}. Our construction
  is an extension of the amortised complexity realisability
  constructions of \cite{hofmann-dallago}. We extend their work to our
  dependently typed setting, and also give a realisability
  interpretation of datatypes directly, instead of via second-order
  impredicative encodings. The technical content of this section has
  been formalised in the Agda proof assistant \cite{agda}, and the
  development is available in the supplementary material.
\end{enumerate}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linear $\lambda$-Calculus and Polytime}
\label{sec:linear-intro}

@@

With booleans and lists, we can construct several other useful
types. For example, to simulate Turing machines, one can construct a
$\mathit{Tape}$ type as a Zipper \cite{huet}
$\ListTy(\BoolTy) \otimes \BoolTy \otimes \ListTy(\BoolTy)$,
representing a position on the tape with the items before, under, and
after the head.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\rho = 0 \lor \sigma = 0)}
  {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
\end{mathpar}
The side conditions on the elimination rule state that (i) both
$\Gamma_1$ and $\Gamma_2$ erase to the same context, so their sum is
defined; and (ii) the argument $N$ is erased (i.e., $\sigma' = 0$) iff
either the function does not use its argument, or we are in the
$\sigma = 0$ fragment and everything is being erased.

@@

  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
\end{mathpar}
Hence, $\Sigma$-types act as normal in the $\sigma = 0$ fragment, and
we can use them as for normal type-theoretic constructions.  In the
$\sigma = 1$ fragment, we must take into account the resource content
of objects and use a pattern matching construct; the dependently typed
analogue of the $\otimes$-eliminator in \autoref{sec:linear-intro}:
\begin{mathpar}
  \inferrule*
  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
\end{mathpar}

@@

The dependently typed recursor for LFPL-style natural numbers again
augments the simply typed linear recursor from
\autoref{sec:lfpl-intro} with dependency information:
\begin{mathpar}
  \mprset{flushleft}
  \inferrule*
  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
    \Gamma \vdash M \stackrel\sigma: \tyNat
    0\Gamma, d \stackrel1: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
    0\Gamma, d \stackrel1: \Diamond, n \stackrel0: \tyNat, p \stackrel1: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x]}
  {\Gamma \vdash \tmRec\,M\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
\end{mathpar}
We have used $* : \Diamond$ as the value in the types for the zero and
successor cases. By the $\eta$-rule for diamonds, we could have
equally well used the $d$ variable that is in scope in each case.

@@

@@

We have only defined an iterable natural number datatype for both of
our systems above. Using the power of dependent types with a universe
we can use these natural numbers to create further datatypes whose
size is measured by a natural. Iteration on the size yields iteration
over the full datastructure. For example, in the LFPL system, we can
define a type of iterable lists by pairing a size with a type of
elements defined by recursion:
\begin{displaymath}
  \mathrm{IList}(A) = \Sigma n \stackrel1: \tyNat.\,\mathrm{rec}_{x.\mathsf{U}}\,n\,\{ \conZero(d) \mapsto I; \conSucc(d,n;p) \mapsto A \otimes p \}
\end{displaymath}

Nil and Cons constructors.

List iterator:

\bob{U notation for the universe above}

\begin{enumerate}
\item Insertion sort in LFPL.
\item Insertion sort in LFPL with correctness specification.
\end{enumerate}

@@

Uses

\begin{displaymath}
  (A : \mathsf{U}, P : A \to \mathsf{U})
\end{displaymath}

Cartesian container morphisms

@@

Monads

@@

@@

@@

\paragraph{Agda Formalisation} The key soundness results in this
section have been formalised in the Agda proof assistant
\cite{agda}. The Agda formalisation can be found in the supplementary
material. After each definition and result we provide a pointer to the
Agda modules where the corresponding formalisation can be found, and
note interesting features of the mechanisation.

@@

\paragraph{Agda Formalisation} The indexed linear preorders are
defined in the Agda module \AgdaModule{IndexedLinear}. We have not yet
completed a formalisation of the construction of a full model of QTT
from an indexed linear preorder, nor have we constructed the syntax of
QTT in Agda, so these parts are currently unmechanised.

@@

Equipped with our underlying costed model of computation
(\autoref{sec:machine-model}) and a compositional notion of resource
potential (\autoref{sec:resource-monoids}), we can construct models of
QTT that witness the resource and type soundness of our complexity
constrained systems. We fix a resource monoid $M$ with distinguished
sub-monoid $M_0$.

@@

We have presented an two extensions of Quantitative Type Theory that
soundly and completely capture polynomial time. We have seen that this
allows for an expressive combination of verification and complexity
constrained computation. We now discuss some related work in implicit
computational complexity and linear dependent types, and take a look
at where the combination of resource conciousness and dependent types
could take us.

@@

The ones from Fossacs 2022, and POPL23(?) Romain Per...

@@

\begin{enumerate}
\item RAML, Hofmann-Jost, Raja, etc.
\item lambda-amor,
\item calf from CMU
\end{enumerate}

@@

\subsection{Towards Synthetic Computational Complexity Theory?}

>>>>

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
\citestyle{acmauthoryear}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Robert Atkey}
\email{robert.atkey@strath.ac.uk}
\orcid{0000-0002-4414-5047}
\affiliation{%
  \institution{University of Strathclyde}
  \streetaddress{26 Richmond Street}
  \city{Glasgow}
  \country{UK}
  \postcode{G1 1XH}
}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We combine dependent types with linear type systems that soundly and
  completely capture polynomial time computation. To combine dependent
  and linear types, we use \emph{Quantitative Type Theory}. We explore
  two systems for capturing polynomial time: one system that disallows
  construction of iterable data, and one, based on the LFPL system of
  Martin Hofmann, that controls construction via a payment
  method. Both of these are extend to full dependent types, allowing
  for arbitrary computation in types alongside guaranteed polynomial
  time computation in terms. We prove the soundness of the systems
  using a realisability technique due to Dal Lago and Hofmann.

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{type theory, implicit computational complexity, linear logic}

@@

Typical Type Theory only speaks of the public face that programs
present to the world -- if you input things like this, you get things
that look like that -- but cannot bring itself to mention the true
cost of programs' execution. One can encode costs by embedding another
programming language in type theory, for example \cite{GueneauCP18},
or one can synthesise costs by treating resource counting as a
computational effect, for example \cite{Danielsson08,NiuSGH22}, but
neither of these capture the intrinsic costs of the programs we write
in type theory. These techniques deliver only \emph{conspicuous
  consumption}, not speaking of the real resources consumed.

@@

This paper makes the following contributions to the theory and use of
linear dependent type theory and implicit computational complexity:
\begin{enumerate}
\item We formulate two systems that combine linear type theory for
  polytime computation with full dependent types, using Quantitative
  Type Theory. The system is presented in \autoref{sec:qtt}. The
  linear typing discipline required for enforcing polytime is provided
  by QTT, but we also need to carefully add constructs for
  non-iterable datatypes (\autoref{sec:noniter-qtt}) and the two kinds
  of natural number iterator that we consider
  (\autoref{sec:cons-free-qtt} and \autoref{sec:lfpl-qtt}). Porting
  the natural number iterators from the simply typed to the
  dependently typed setting requires careful annotation of the rules
  to ensure that the correct information is available for type
  checking, while also not allowing too much information to be made
  available at runtime that would violate the polytime soundness
  property. A further contribution of this paper is the addition of
  reflection types to QTT, \autoref{sec:qtt-reflection}, which allow
  statements about polytime realisability to be reflected into
  types. % We explain in \autoref{Sec:realisability-future} how this
  % will have applications in QTT beyond the polytime systems presented
  % here.
\item We demonstrate the utility of the combination of polytime and
  dependent types in \autoref{sec:programming-polytime}. Just as in
  the simply typed world, we have an expressive language for writing
  polytime programs. With the additional power of dependent types, we
  can also prove properties of these programs. A simple example is
  proving that a sorting program actually sorts. Using QTT reflection,
  we can go further and represent the class of polytime problems, with
  polytime reductions between them, as dependent pairs
  (\autoref{sec:polytime-problems}). Our final examples use dependent
  types to give monadic presentations of the complexity classes of
  Non-deterministic Polynomial (NP) time and Probabilistic Polynomial
  (PP) time. Since these classes rely on specific correctness
  criteria, it is not possible to capture them in a simply typed
  system for polytime.
\item We prove the polytime soundness of our systems via a
  realisability argument in \autoref{sec:soundness}. Our construction
  is an extension of the amortised complexity realisability
  constructions of \citet{dallago11realisability}. We extend their
  work to our dependently typed setting, and also give a realisability
  interpretation of datatypes directly, instead of via second-order
  impredicative encodings. The technical content of this section has
  been formalised in the Agda proof assistant
  \cite{norell2008dependently}, and is included in the supplementary
  material.
\end{enumerate}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Affine Linear $\lambda$-Calculus and Polytime}
\label{sec:linear-intro}

@@

With booleans and lists, we can construct several other useful
types. For example, to simulate Turing machines, one can construct a
$\mathrm{Tape}$ type as a Zipper \cite{huet}
$\ListTy(\BoolTy) \otimes \BoolTy \otimes \ListTy(\BoolTy)$,
representing a position on the tape with the items before, under, and
after the head.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\rho = 0 \lor \sigma = 0)}
  {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
\end{mathpar}
The side conditions on the elimination rule state that (i) both
$\Gamma_1$ and $\Gamma_2$ erase to the same context, so their sum is
defined; and (ii) the argument $N$ is erased (i.e., $\sigma' = 0$) iff
either the function does not use its argument, or we are in the
$\sigma = 0$ fragment and everything is being erased. In the
following, when we write $S \to T$ for a non-dependent function type,
we mean that the argument is to be used linearly:
$(x \stackrel1: S) \to T$.

@@

  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
\end{mathpar}
Hence, $\Sigma$-types act as normal in the $\sigma = 0$ fragment, and
we can use them as for normal type-theoretic constructions.  In the
$\sigma = 1$ fragment, we must take into account the resource content
of objects and use a pattern matching construct; the dependently typed
analogue of the $\otimes$-eliminator in \autoref{sec:linear-intro}:
\begin{mathpar}
  \inferrule*
  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
\end{mathpar}
QTT also supports a unit type $I$ with constructor $*$ and
pattern-matching elimination. See \cite{atkey18qtt} for details.

@@

The dependently typed recursor for LFPL-style natural numbers again
augments the simply typed linear recursor from
\autoref{sec:lfpl-intro} with dependency information:
\begin{mathpar}
  \mprset{flushleft}
  \inferrule*
  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
    \Gamma \vdash M \stackrel\sigma: \tyNat \\\\
    0\Gamma, d \stackrel\sigma: \Diamond \vdash N_z \stackrel\sigma: P[\conZero(*)/x] \\\\
    0\Gamma, d \stackrel\sigma: \Diamond, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma : P[\conSucc(*,n)/x]}
  {\Gamma \vdash \tmRec\,M\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
\end{mathpar}
We have used $* : \Diamond$ as the value in the types for the zero and
successor cases. By the $\eta$-rule for diamonds, we could have
equally well used the $d$ variable that is in scope in each case.

@@

We now explore the possibilities afforded by the combination of
polytime guarantees with the specification expressivity of dependent
types.

@@

We have only defined an iterable natural number datatype for both of
our systems above. We could extend both systems to include further
iterable inductive types, although in the \emph{Cons-free} system this
is not particularly useful due to the prohibition of on
construction. However, sticking with just the natural numbers, we can
use the power of dependent types with a universe to create further
datatypes whose size is measured by some iterable natural
number. Iteration on the size yields iteration over the full
datastructure. For example, in the LFPL system, we can define a type
of iterable lists by pairing a size with a type of elements defined by
recursion on the size:
\begin{displaymath}
  \mathrm{IList}\,A = (n \stackrel1: \tyNat) \otimes \left(\mathrm{rec}_{x.\mathsf{U}}\,n\,\{ \conZero(d) \mapsto I; \conSucc(d,n;p) \mapsto A \otimes p \}\right)
\end{displaymath}
The $\cNil$ and $\cCons$ constructors can now be defined in terms of
$\conZero$ and $\conSucc$, provided the caller supplies sufficient
$\Diamond$s. These definitions live in the $\sigma = 1$ fragment, so
we annotate them appropriately:
\begin{mathpar}
  \begin{array}{l}
    \mathrm{nil} \stackrel1: \Diamond \to \mathrm{IList}\,A \\
    \mathrm{nil}\,d = (\conZero(d), *)
  \end{array}

  \begin{array}{l}
    \mathrm{cons} \stackrel1: \Diamond \to A \to \mathrm{IList}\,A \to \mathrm{IList}\,A \\
    \mathrm{cons}\,d\,x\,\mathit{xs} = \mathrm{let}\,(n,\mathit{elems}) = \mathit{xs}\,\textrm{in}\,(\conSucc(d,n),(x,\mathit{elems}))
  \end{array}
\end{mathpar}
Using the LFPL iterator it is also possible to construct a dependently
typed iterator for $\mathrm{IList}\,A$ values. Unfortunately, the
current types of the LFPL system are not sufficient to type this as a
function, as we have no way of stating that the successor case must be
arbitrarily duplicable. Lifting this restrict by means of some
modality is future work. The typing rule for the derived list iterator
is:
\begin{mathpar}
  \mprset{flushleft}
  \inferrule*
  {0\Gamma \vdash A~\istype \\
    0\Gamma, x \stackrel0: \mathrm{IList}\,A \vdash P~\istype \\\\
    \Gamma \vdash M \stackrel\sigma: \mathrm{IList}\,A \\\\
    0\Gamma, d \stackrel1: \Diamond \vdash N_1 \stackrel\sigma: P[\mathrm{nil}(*)/x] \\\\
    0\Gamma, d \stackrel1: \Diamond, x \stackrel\sigma: A, \mathrm{xs} \stackrel0: \mathrm{IList}\,A, p \stackrel\sigma: P[\mathit{xs}/x] \vdash N_2 \stackrel\sigma : P[\mathrm{cons}(*,x,\mathit{xs})/x]}
  {\Gamma \vdash \tmRec_{x.P}\,M\,\{\mathrm{nil}(d) \mapsto N_1; \mathrm{cons}(d,x,\mathit{xs};p) \mapsto N_2\} \stackrel\sigma: P[M/x]}
\end{mathpar}
Note that, in the cons case, we have access to the result of iterating
over the tail of the list ($p$), but not to the actual tail of the
list ($\mathit{xs}$).

With our list iterator, it is now possible to write interesting
polytime programs. For example, insertion of a natural into a sorted
list:
\begin{displaymath}
  \mathrm{insert} \stackrel1: \Diamond \to \tyNat \to \mathrm{IList}\,\tyNat \to \mathrm{IList}\,\tyNat
\end{displaymath}
which requires some ingenuity to write to handle the case where we
find the place to insert the item and need access to the remainder of
the list. Note that, also, the function consumes a $\Diamond$ to
construct the new element of the output list, and also that the items
in the list are themselves iterable natural numbers. This is needed to
account for the

Insertion sort is repeated insertion of elements from an original list
into a new list. The new list is constructed from the $\Diamond$s
yielded by the original list:
\begin{displaymath}
  \mathrm{insertionSort} \stackrel1: \mathrm{IList}\,A \to \mathrm{IList}\,A
\end{displaymath}
The immediate benefit of dependent types in this situation is that it
is now possible to state and prove the correctness property of this
sorting procedure. Using the fact that the $\sigma = 0$ fragment of
QTT is exactly normal type theory, we can use normal dependently typed
programmming techniques to establish:
\begin{displaymath}
  \mathrm{insertionSortCorrect} \stackrel0: (\mathit{xs} \stackrel1: \mathrm{IList}\,A) \to \mathrm{Sorted}(\mathit{xs},\mathrm{insertionSort}\,\mathrm{xs})
\end{displaymath}
where $\mathrm{Sorted}(x,y)$ is some predicate stating that $y$ is a
sorted permutation of $x$.  Note that, despite the $1$ annotation on
the $\Pi$-type here, we are free to duplicate $\mathit{xs}$ because
types are constructed in the $\sigma = 0$ fragment.

@@

Define a decision problem to be a pair $(A, P)$, where $A$ is a type
in the universe $\mathsf{U}$, and $P : A \to \mathsf{U}$ is a
predicate on $A$. For what follows, we are only interested in whether
or not $P\,a$ is inhabited for each $a$. Therefore, we use
$P \Leftrightarrow Q$ to stand for equi-inhabitation of two
$P, Q : \mathsf{U}$, i.e.,
$P \Leftrightarrow Q \equiv (P \to Q) \times (Q \to P)$.

We can use the reflection type former defined in
\autoref{sec:qtt-reflection} to define a predicate on decision
problems that establishes whether or not they are polytime decision
problems. Specifically, we can state that there is a polytime
realisable boolean-value predicate that reports true exactly when the
given element of $a$ is in the predicate:
\begin{displaymath}
  \mathrm{PTIME}(A,P) =
    (f \stackrel1: \mathbf{R}(A \to \BoolTy)) \otimes (a \stackrel1: A) \to (\mathbf{R}^{-1}(f)\, a = \cTrue) \Leftrightarrow P\,a
\end{displaymath}
Thus, $\mathrm{PTIME}(A,P)$ is a logical proposition stating that the
decision problem $(A,P)$ is deciable in polytime. We make two notes
about this definition: (i) this definition is intrinsic, in the sense
that, whichever of the polytime systems is chosen, proving that a
decision problem is solvable in polytime is a matter of programming,
without having to reason directly about machine models and step
counting; and (ii) we have defined problems to have arbitrary types
$A$ as domains, rather than bitstrings, and so the notion of size
attached to an input is intrinsic to the type $A$ chosen.

We can also declare a type of polytime \emph{reductions} between
problems. A problem $(A, P)$ can be polytime reduced to a problem
$(B,Q)$ if there is an inhabitant of the following type:
\begin{displaymath}
  (A,P) \stackrel{\textsc{Poly}}\Rightarrow (B,Q) = (f \stackrel1: \mathbf{R}(A \to B)) \otimes (a \stackrel1: A) \to Q(\mathbf{R}^{-1}(f)\,a) \Leftrightarrow P\,a
\end{displaymath}
In words, there must be a polytime function $f$ that preserves and
reflects decisions. With this definition, it is possible to prove in
our systems that polytime computations are closed under polytime
reductions. We note that this definition is, up to the reflection
modality, the same as the definition of cartesian container morphism,
well known in dependent type theory \cite{abbott03categories}, and
speaks to a general conception of containers as ``problem/solution''
pairings.

@@

The fact that we can characterise polytime decision problems is
perhaps to be expected from a system designed to capture polynomial
time realisable programs. However, we can go further to capture the
complexity classes of Non-deterministic Polynomial time (NP) and
Probabilistic Polynomial time (PP), both of which are based on
polytime. We do this by augmenting our polytime functions with
additional power in the form of computation effects.

@@

To capture the complexity class NP, we use polynomial time programs
augmented with non-determinism, as one might expect. We will not need
to reason about equality of these non-deterministic programs, so we
can represent non-deterministic choices as \emph{additive} binary
trees. We suppose a non-iterable datatype defined like so:
\begin{displaymath}
  \begin{array}{l}
    \textbf{data}\,\mathrm{ND}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
    \quad \mathsf{return} : A \to \mathrm{ND}\,A \\
    \quad \mathsf{choice} : (\BoolTy \to \mathrm{ND}\,A) \to \mathrm{ND}\,A
  \end{array}
\end{displaymath}
The crucial point here is that the subtrees are represented as a
function $\BoolTy \to \mathrm{ND}\,A$. By the typing rules of QTT,
this means that the two branches of this function can share resources
(see the encoding of the additive product types in
\cite{atkey18qtt}). Thus, each branch of this tree is polynomial time,
not the whole tree itself.

The type $\mathrm{ND}$ supports a monad interface via the usual free
monad construction, as well as an effect
$\mathrm{flip} \stackrel1: \mathrm{ND}\,\BoolTy$ providing access to a
bit of non-deterministic information. Thus a program of type
$A \to \mathrm{ND}\,B$ in the $\sigma = 1$ fragment will be a polytime
program with access to an oracle. In the $\sigma = 0$ fragment, we can
write a function that resolves non-determinism using a list of
booleans. This function returns $\mathsf{nothing}$ if the list of
booleans is insufficient to resolve all the $\mathsf{choice}$s:
\begin{displaymath}
  \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A
\end{displaymath}
With these definitions, we can define Non-deterministic Polynomial
time as a predicate on problems:
\begin{displaymath}
  \mathrm{NP}(A,P) =
  \begin{array}[t]{l}
    (f \stackrel1: \mathbf{R}(A \multimap \mathrm{ND}(\BoolTy)))\, \otimes \\
    \quad (a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a
  \end{array}
\end{displaymath}
Thus, a problem is in NP if there is a non-deterministic
boolean-valued polynomial time function that has a path to returning
$\cTrue$ exactly when the input is in predicate. Moreover, it is a
quick matter of programming to see that problems in NP are closed
under the type of polytime reductions given above.

% \begin{displaymath}
%   \begin{array}{l}
%     \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A \\
%     \mathrm{runWithOracle}\,(\mathsf{return}\,a)\,\mathit{bs} = \mathsf{just}\,a \\
%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,(\cTrue :: \mathit{bs}) = \mathrm{runWithOracle}\,x\,\mathit{bs} \\
%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,(\cFalse :: \mathit{bs}) = \mathrm{runWithOracle}\,y\,\mathit{bs} \\
%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,\cNil = \mathsf{nothing}
%   \end{array}
% \end{displaymath}

@@

By changing the computation effects supplied to a program, we can
change the complexity class. To capture Probabilistic Polynomial time,
we use a (non-iterable) data structure representing trees of
probabilistic choices, where $\mathbb{Q}[0,1]$ is some type of
(non-iterable) rationals in the closed interval $[0,1]$:
\begin{displaymath}
  \begin{array}{l}
    \textbf{data}\,\mathrm{Dist}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
    \quad \mathsf{return} : A \to \mathrm{Dist}\,A \\
    \quad \mathsf{choice} : \mathbb{Q}[0,1] \to (\BoolTy\ to \mathrm{Dist}\,A) \to \mathrm{Dist}\,A
  \end{array}
\end{displaymath}
As in the non-deterministic case, a function $A \to \mathrm{Dist}\,B$
in the $\sigma = 1$ fragment is a polytime probabilistic
computation. Again, the use of a function type here ensures that each
branch of the tree is constructable in polynomial time, not the whole
tree. In the $\sigma = 0$ fragment we can write a function that
computes the probability of a $\mathrm{Dist}\,\BoolTy$ computation
being true:
\begin{displaymath}
  \mathrm{probTrue} \stackrel0: \mathrm{Dist}\,\BoolTy \to \mathbb{Q}[0,1]
\end{displaymath}
We can now define the class of probabilistic polynomial time decision
problems, where the decider is allowed to make probabilistic choices
as long as it is correct with probability at least $\frac{1}{3}$:
\begin{displaymath}
  \mathrm{PP}(A,P) =
  \begin{array}[t]{l}
    (f \stackrel1: \mathbf{R}(A \multimap \mathrm{Dist}(\BoolTy))) \, \otimes
    (a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{1}{3}) \Leftrightarrow P\,a
  \end{array}
\end{displaymath}
Again, problems in PP are easily seen to be closed under polytime
reductions.

@@

\paragraph{Agda Formalisation} The key soundness results in this
section have been formalised in the Agda proof assistant
\cite{norell2008dependently}. The Agda formalisation can be found in
the supplementary material. After each definition and result we
provide a pointer to the Agda modules where the corresponding
formalisation can be found, and note interesting features of the
mechanisation.

@@

\paragraph{Agda Formalisation} The indexed linear preorders are
defined in the Agda module \AgdaModule{IndexedLinear}. We have not yet
completed a formalisation of the construction of a full model of QTT
from an indexed linear preorder so this part is currently unmechanised.

@@

Equipped with our underlying costed model of computation
(\autoref{sec:machine-model}) and a compositional notion of resource
potential (\autoref{sec:resource-monoids}), we can construct models of
QTT that witness the resource and type soundness of our complexity
constrained systems. We fix a resource monoid $M$ with sub-monoid
$M_0$ and proceed to build an indexed linear preorder of resource
accounted realisers.

@@

We have presented two extensions of Quantitative Type Theory that
soundly and completely capture polynomial time. We have seen that this
allows for an expressive combination of verification and complexity
constrained computation. We now discuss some related work in implicit
computational complexity and linear dependent types, and take a look
at where the combination of resource conciousness and dependent types
could take us.

@@

The ones from Fossacs 2022, and POPL23(?) Romain Perchoux

@@

\begin{enumerate}
\item RAML, Hofmann-Jost, Raja, etc.
\item lambda-amor,
\item Daniellson
\item calf from CMU
\end{enumerate}

@@

\subsection{Towards Synthetic Computational Complexity Theory?}
\label{sec:towards-synthetic-complexity-theory}
